A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-04 tex/codatadp.tex
more

@@

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$), and it is treated as a logical symbol,
just like the connectives and quantifiers of first-order logic.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
for each i, j: (\forall \bar x. IH_{ij}[\bar x] --> P_i(C_{ij}(\bar x)))
---------------------------
P_1[x_1] \& ... \& P_l[x_l]
\end{aligned}
\]
%
Ensures that we get the smallest model. E.g. for natural numbers, this rules out
nonstandard models, which could contain cyclic elements (e.g., n such that
n = Suc n) or even infinite acyclic elements $Suc (Suc \ldots)$.

For codatatype, the dual notion is that of a coinduction principle that
depends on witnesses $R$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Conduction:}\quad
for each i: R_i[x_i, y_i]
for each i: (\forall x y. R_i[x, y] --> \bigwedge_k s_{ij}^k(x) \sim s_{ij}^k(y))
---------------------------------
x_1 \teq y_1 & ... & x_l \teq y_l
\end{aligned}
\]

>>>>

% for "bussproofs" package
\EnableBpAbbreviations
\def\ScoreOverhang{1.5pt}
\def\proofSkipAmount{\vskip 0pt}
\def\defaultHypSeparation{\hskip0.75em}

@@

The proofs are part of a report.

@@

\paragraph{Conventions.}
This paper assumes a monomorphic (or many-sorted) first-order logic throughout.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are simply atomic sorts, with no
structure, and interpreted by nonempty domains. The set $\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values $\{\bot, \top\}$. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The only predicate is equality ($\teq$), and it is treated as a logical symbol,
just like the connectives and quantifiers of first-order logic.
Other predicates can be represented as functions that return $\ty{bool}$.
The metavariables $\delta, \tau$ range over types, with $\delta$ reserved
for datatypes and codatatypes.

@@

Datatypes are additionally characterized by an induction axiom schema:
%
\[
\begin{aligned}[t]
\text{Induction:}\quad
\AXC{$\forall x_1 \ldots \vthinspace x_{n_{ij}}.\; \bigl(\bigwedge_{\,k\,} \mathit{IH}_{ij}^k[\bar x_k]\bigr) \longrightarrow P_i[C_{ij}(\bar x)]$\quad\text{for each $i,j$}}
\UIC{$P_1[v_1] \mathrel\land \cdots \mathrel\land P_l[v_l]$}
\DP
\end{aligned}
\]
where the induction hypothesis $\mathit{IH}_{ij}^k(x)$
denotes either $P_{i'}(x)$ for some $i$ such that this is type
correct or else $\top$.
%
Ensures that we get the smallest model. E.g. for natural numbers, this rules out
nonstandard models, which could contain cyclic elements (e.g., n such that
n = Suc n) or even infinite acyclic elements $Suc (Suc \ldots)$.

For codatatype, the dual notion is that of a coinduction principle that
depends on witnesses $R$ that are required to be bisimulations:
%
\[
\begin{aligned}[t]
\text{Conduction:}\quad
\AXC{for each i: $R_i[x_i, y_i]$}
\AXC{for each i: $(\forall x y. R_i[x, y] \longrightarrow \bigwedge_k s_{ij}^k(x) \sim s_{ij}^k(y))$}
\BIC{$x_1 \teq y_1 \& ... \& x_l \teq y_l$}
\DP
\end{aligned}
\]
